---
title: "Homophily as a process generating social networks"
description: "R code of the analysis."
author:
  - name: Szymon Talaga 
    url: https://github.com/sztal
    affiliation: The Robert Zajonc Institute for Social Studies, University of Warsaw
    affiliation_url: https://iss.uw.edu.pl
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup_markdown, include=FALSE}
PACKAGES <- c("here", "Cairo")
for (p in PACKAGES) {
    if (!(p %in% installed.packages())) {
        install.packages(p)
    }
}
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 8,
    fig.asp = 0.618,
    base.dir = here::here(),
    root.dir = here::here(),
    fig.path = "figures/",
    dev = "CairoPS",
    dev.args = list(
        family = "sans"
    ),
    cache.path = ".cache/knitr/"
)
```


```{r setup}
library(magrittr)
library(boot)
library(tidyverse)
library(feather)
library(emmeans)
library(broom)
library(ggsignif)
library(ggpubr)
library(gridExtra)
library(DescTools)


# Setup ggplot aesthetics
theme_set(theme_bw(
    # base_family = "sans",
    base_size = 13
))

# Globals
HERE <- here::here()
ROOT <- file.path(HERE, "..")
DATA <- file.path(ROOT, "data")

source(file.path(HERE, "functions.R"))

SIZES <- c(1000, 2000, 4000, 8000)
ALPHA_COLORS <- c("#5780e0", "#7db717", "#d37e00", "#d80000")
ALPHA_LABELS <- c("2", "4", "8", bquote("\U221E"))
LEGEND_TR <- c(.834, .945)
LEGEND_TL <- c(.163, .945)

# Set random seed
set.seed(1010)
```

# SDA model simulations

```{r sda_get_data}
sda <- read_feather(file.path(DATA, "sda-data.feather")) %>%
    mutate(
        degseq = map(degseq, ~as.integer(str_split(.x, pattern = fixed("|"))[[1]])),
        space = space %>% fct_recode(
            Uniform = "uniform",
            `Log-normal` = "lognormal",
            `Gaussian clusters` = "clusters_normal"
        ) %>% fct_relevel("Uniform", "Gaussian clusters", "Log-normal"),
        alpha_f = factor(alpha, levels = c(2, 4, 8, Inf)),
        m_f = factor(str_c("m = ", m), levels = c("m = 1", "m = 2", "m = 4", "m = 8", "m = 16")),
        p_rewire_f = if_else(p_rewire == 0, "No rewiring", "Random rewiring") %>%
            factor(levels = c("No rewiring", "Random rewiring")),
        density = k_avg / (N - 1),
        clustering_c = clustering - density,
        alpha_n = if_else(is.finite(alpha), alpha, 100)
    ) %>%
    mutate(deg_gini = map_dbl(degseq, Gini, unbiased = TRUE))
```

## Clustering

```{r sda_clustering, fig.width = 8, fig.asp = 0.618}
gdf <- sda %>%
    group_by(space, N, m_f, alpha_f) %>%
    summarize(
        stat = mean(clustering),
        lower = simple_boot(clustering, min),
        upper = simple_boot(clustering, max),
        n = n()
    ) %>%
    ungroup

gdf %>%
    make_line_plot(LEGEND_TR) +
    facet_grid(space ~ m_f) +
    scale_y_continuous(limits = c(0, 1)) +
    ylab("Clustering")
```

## Degree assortativity

```{r sda_assortativity}
gdf <- sda %>%
    group_by(space, N, m_f, alpha_f) %>%
    summarize(
        stat = mean(assortativity),
        lower = simple_boot(assortativity, min),
        upper = simple_boot(assortativity, max),
        n = n()
    ) %>%
    ungroup

gdf %>%
    make_line_plot(LEGEND_TR, hline = 0) +
    facet_grid(space ~ m_f) +
    scale_y_continuous(limits = c(-.2, 1), breaks = c(0, .25, .5, .75, 1)) +
    ylab("Degree assortativity")
```

## Average path lengths (small-world)

```{r sda_average_path_lengths}
sda %>%
    group_by(p_rewire, m_f, alpha_f, space) %>%
    summarize(
        r = cor(average_path_length, log(N), method = "pearson")
    ) %>%
    ungroup %>%
    split(.$p_rewire) %>%
    map(summary)

gdf <- sda %>%
    group_by(p_rewire_f, N, m_f, alpha_f) %>%
    summarize(
        stat = mean(average_path_length),
        lower = simple_boot(average_path_length, min),
        upper = simple_boot(average_path_length, max),
        n = n()
    ) %>%
    ungroup

gdf %>%
    make_line_plot(LEGEND_TR) +
    facet_grid(p_rewire_f ~ m_f, scales = "free_y") +
    ylab("Average path length")

gdf %>%
    filter(p_rewire_f == "No rewiring") %>%
    make_line_plot("top") +
    facet_wrap(~m_f, scales = "free_y") +
    ylab("Average path length")
```

```{r sda_average_path_length_correlations}
Q <- sda %>%
    group_by(space, m, alpha, p_rewire) %>%
    summarize(
        L_cor = cor(average_path_length, log(N))
    )
small_world <- tapply(Q$L_cor, Q$p_rewire, summary)
```

## Gini coefficient

```{r sda_gini, fig.width = 8, fig.asp = .618}
gdf <- sda %>%
    group_by(space, N, m_f, alpha_f) %>%
    summarize(
        stat = mean(deg_gini),
        lower = simple_boot(deg_gini, min),
        upper = simple_boot(deg_gini, max),
        n = n()
    ) %>%
    ungroup

gdf %>%
    make_line_plot(LEGEND_TL, hline = 0) +
    facet_grid(space ~ m_f) +
    ylab("Gini coefficient")
```

## Degree distributions

```{r sda_deg_dists, fig.width = 8, fig.asp = .618}
gdf <- sda %>% 
    filter(N == 8000) %>%
    group_by(space, N, m_f, alpha_f) %>% 
    summarize(deg = list(reduce(degseq, c))) %>% 
    unnest %>%
    ungroup

{ gdf %>%
    ggplot(aes(x = deg, fill = alpha_f)) +
    geom_density(alpha = .7) +
    facet_grid(space ~ m_f, scales = "free_x") +
    scale_fill_manual(values = ALPHA_COLORS, labels = ALPHA_LABELS) +
    labs(x = "Node degree", y = "Probability density", fill = "α") +
    theme(
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5),
        legend.position = c(.835, .945),
        legend.background = element_rect(fill = "white", color = "black"),
        legend.text = element_text(size = 9, face = "bold"),
        legend.title.align = .5,
        legend.direction = "horizontal"
    )
} %>%
    add_base_theme()
```

## Statistical models

### Statistical model for clustering

```{r sda_models_clustering}
cl_lm <- lm(logit(clustering) ~ space + log(N) + log(m) + log(alpha_n), data = sda)
summary(cl_lm)
cl_emm <- emmeans(cl_lm, ~space, type = "response")
cl_cnt <- contrast(cl_emm, method = "consec") %>% tidy
cl_plt <- (cl_emm %>%
    tidy %>%
    ggplot(aes(x = space, y = response, color = space)) +
    geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = .25) +
    geom_point(shape = 16, size = 3) +
    geom_signif(
        color = "black",
        annotations = "< 0.001", 
        xmin = 1, 
        xmax = 2,
        y_position = .35,
        tip_length = c(1, .25)
    ) +
    geom_signif(
        color = "black",
        annotations = "< 0.001",
        xmin = 2,
        xmax = 3,
        y_position = .35,
        tip_length = c(0, .2)
    ) +
    scale_color_manual(values = ALPHA_COLORS) +
    scale_y_continuous(limits = c(0, .4)) +
    theme(
        axis.text.x = element_text(size = 9)
    ) +
    guides(color = FALSE)) %>%
    add_base_theme +
    xlab("") +
    ylab("Clustering")
```

### Statistical model for assortativity

```{r sda_models_assortativity}
as_lm <- lm(atanh(assortativity) ~ space + log(N) + log(m) + log(alpha_n), data = sda)
summary(as_lm)
as_emm <- emmeans(as_lm, ~space)
as_cnt <- contrast(as_emm, method = "consec") %>% tidy
as_plt <- (as_emm %>%
    tidy %>%
    mutate_at(vars(estimate, conf.low, conf.high), tanh) %>%
    ggplot(aes(x = space, y = estimate, color = space)) +
    geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = .25) +
    geom_point(shape = 16, size = 3) +
    geom_signif(
        color = "black",
        annotations = "< 0.001", 
        xmin = 1, 
        xmax = 2,
        y_position = .6,
        tip_length = c(.55, .1)
    ) +
    geom_signif(
        color = "black",
        annotations = "< 0.001",
        xmin = 2,
        xmax = 3,
        y_position = .8,
        tip_length = c(.5, .15)
    ) +
    scale_color_manual(values = ALPHA_COLORS) +
    scale_y_continuous(limits = c(0, 1)) +
    theme(
        axis.text.x = element_text(size = 9)
    ) +
    guides(color = FALSE)) %>%
    add_base_theme +
    xlab("") +
    ylab("Degree assortativity")
```

### Model-based differences between types of social space

```{r sda_models_plot, fig.width = 8}
grid.arrange(cl_plt, as_plt, ncol = 2)
```

# SDA configuration model simulation results

```{r sdc_get_data}
sdc <- read_feather(file.path(DATA, "sda-data-cm.feather")) %>%
    mutate(
        degseq = map(degseq, ~as.integer(str_split(.x, pattern = fixed("|"))[[1]])),
        space = space %>% fct_recode(
            Uniform = "uniform",
            `Log-normal` = "lognormal",
            `Gaussian clusters` = "clusters_normal"
        ) %>% fct_relevel("Uniform", "Gaussian clusters", "Log-normal"),
        alpha_f = factor(alpha, levels = c(2, 4, 8, Inf)),
        m_f = factor(str_c("m = ", m), levels = c("m = 1", "m = 2", "m = 4", "m = 8", "m = 16")),
        p_rewire_f = if_else(p_rewire == 0, "No rewiring", "Random rewiring") %>%
            factor(levels = c("No rewiring", "Random rewiring")),
        density = k_avg / (N - 1),
        clustering_c = clustering - density,
        degseq_type = factor(degseq_type) %>% fct_recode(
            Poisson = "poisson",
            `Negative binomial` = "negbinom",
                `Power-law` ="powerlaw"
        ) %>% fct_relevel("Poisson", "Negative binomial", "Power-law"),
        alpha_n = if_else(is.finite(alpha), alpha, 100)
    )
```

## Clustering

```{r sdc_clustering, fig.width = 8, fig.asp = 0.618}
gdf <- sdc %>%
    group_by(degseq_type, N, m_f, alpha_f) %>%
    summarize(
        stat = mean(clustering),
        lower = simple_boot(clustering, min),
        upper = simple_boot(clustering, max),
        n = n()
    ) %>%
    ungroup

gdf %>%
    make_line_plot(legend_position = LEGEND_TR) +
    facet_grid(degseq_type ~ m_f) +
    scale_y_continuous(limits = c(0, 1)) +
    ylab("Clustering")
```

## Degree assortativity

```{r sdc_assortativity}
gdf <- sdc %>%
    group_by(degseq_type, N, m_f, alpha_f) %>%
    summarize(
        stat = mean(assortativity),
        lower = simple_boot(assortativity, min),
        upper = simple_boot(assortativity, max),
        n = n()
    ) %>%
    ungroup

gdf %>%
    make_line_plot(legend_position = LEGEND_TR, hline = 0) +
    facet_grid(degseq_type ~ m_f) +
    scale_y_continuous(limits = c(-.2, 1), breaks = c(0, .25, .5, .75, 1)) +
    ylab("Degree assortativity")
```

## Average path lengths (small-world)

```{r sdc_average_path_lengths}
sdc %>%
    group_by(p_rewire, m_f, alpha_f, space, degseq_type, degseq_sort) %>%
    summarize(
        r = cor(average_path_length, log(N), method = "pearson")
    ) %>%
    ungroup %>%
    split(.$p_rewire) %>%
    map(summary)

gdf <- sdc %>%
    group_by(degseq_type, N, m_f, alpha_f) %>%
    summarize(
        stat = mean(average_path_length),
        lower = simple_boot(average_path_length, min),
        upper = simple_boot(average_path_length, max),
        n = n()
    ) %>%
    ungroup

gdf %>%
    make_line_plot(LEGEND_TR) +
    facet_grid(degseq_type ~ m_f, scales = "free_y") +
    ylab("Average path length")
```

## SDA-SDC comparison

```{r sda_sdc_comparison, fig.width = 8, fig.height = 5}
df1 <- sda %>%
    select(space, N, m, alpha_f, clustering, assortativity) %>%
    group_by(space, N, m, alpha_f) %>%
    summarize(
        clustering = list(clustering),
        assortativity = list(assortativity)
    ) %>%
    ungroup
df2 <- sdc %>%
    select(degseq_type, space, N, m, alpha_f, clustering, assortativity) %>%
    group_by(degseq_type, space, N, m, alpha_f) %>%
    summarize(
        clustering = list(clustering),
        assortativity = list(assortativity)
    ) %>%
    ungroup

df <- left_join(df1, df2, by = c("space", "N", "m", "alpha_f"), suffix = c("", "_cm")) %>%
    mutate(
        clustering = map2(clustering, clustering_cm, ~list(.x, .y)),
        assortativity = map2(assortativity, assortativity_cm, ~list(.x, .y))
    ) %>%
    mutate(
        cl_comparison = map(clustering, ~{
            data.frame(
                cl_delta = median(.x[[1]]) - median(.x[[2]]),
                cl_p_val = wilcox.test(.x[[1]], .x[[2]], exact = FALSE)$p.value
            )
        }),
        as_comparison = map(assortativity, ~{
            data.frame(
                as_delta = median(.x[[1]] - median(.x[[2]])),
                as_p_val = wilcox.test(.x[[1]], .x[[2]], exact = FALSE)$p.value
            )
        })
    ) %>%
    select(-clustering, -clustering_cm, -assortativity, -assortativity_cm) %>%
    unnest(cl_comparison, as_comparison) %>%
    mutate_at(vars(cl_p_val, as_p_val), p.adjust, method = "BY")

plt1 <- (df %>%
    mutate(cl_delta = if_else(cl_p_val > 0.01, 0, cl_delta)) %>%
    ggplot(aes(x = alpha_f, y = as.factor(m), fill = cl_delta)) +
    geom_tile() +
    facet_grid(degseq_type ~ space) +
    scale_fill_gradient(low = "#ffffff", high = ALPHA_COLORS[2], name = "Me(SDA) - Me(SDC)")
) %>%
    add_base_theme +
    theme(
        legend.position = "top",
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 7),
        strip.text = element_text(size = 7),
        axis.title = element_text(size = 9)
    ) +
    xlab("Homophily (α)") +
    ylab("Number of dimensions (m)") +
    ggtitle("Clustering")

plt2 <- (df %>%
    mutate(as_delta = if_else(cl_p_val > 0.01, 0, as_delta)) %>%
    ggplot(aes(x = alpha_f, y = as.factor(m), fill = as_delta)) +
    geom_tile() +
    facet_grid(degseq_type ~ space) +
    scale_fill_gradient(low = "#ffffff", high = ALPHA_COLORS[2], name = "Me(SDA) - Me(SDC)")
) %>%
    add_base_theme +
    theme(
        legend.position = "top",
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 7),
        strip.text = element_text(size = 7),
        axis.title = element_text(size = 9),
    ) +
    xlab("Homophily (α)") +
    ylab("Number of dimensions (m)") +
    ggtitle("Degree assortativity")

ggarrange(plt1, plt2, nrow = 1, ncol = 2, common.legend = TRUE, legend = "top")
```

## Statistical models

### Statistical model for clustering

```{r sdc_models_clustering}
cl_cm_lm <- lm(logit(clustering) ~ space * degseq_type + log(N) + log(m) + log(alpha_n), data = sdc)
summary(cl_cm_lm)
cl_cm_emm <- emmeans(cl_cm_lm, ~space + degseq_type, type = "response")
cl_cm_cnt <- contrast(cl_cm_emm, method = "pairwise", by = "degseq_type")

sig_df <- tibble(
    degseq_type = c(rep("Poisson", 2), rep("Negative binomial", 2), "Power-law"),
    x = c(1.5, 2.5, 1.5, 2.5, 2.5),
    y = c(.086, .093, .137, .14, .17),
    label = c("< 0.001", "0.001", "< 0.001", "< 0.001", "0.045")
)

cl_cm_plt <- (cl_cm_emm %>%
    tidy %>%
    ggplot(aes(x = space, y = response, color = degseq_type, group = degseq_type)) +
    geom_line(linetype = 2) +
    geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = .25) +
    geom_point(shape = 16, size = 2) +
    geom_label(data = sig_df, aes(x = x, y = y, label = label, color = degseq_type), size = 3) +
    geom_signif(
        color = "black",
        annotations = "< 0.001",
        xmin = 1,
        xmax = 3,
        y_position = .05,
        tip_length = c(-.1, -.2),
        vjust = 4,
        textsize = 3
    ) +
    scale_color_manual(values = ALPHA_COLORS, name = "Degree sequence") +
    scale_y_continuous(limits = c(0, .2), breaks = c(0, .05, .1, .15, .2)) +
    theme(
        axis.text.x = element_text(size = 9),
        legend.background = element_rect(fill = "white", color = "black"),
        legend.position = c(.12, .1),
        legend.title = element_text(size = 9, face = "bold"),
        legend.text = element_text(size = 9)
    )
) %>%
    add_base_theme +
    xlab("") +
    ylab("Clustering")
```

### Statistical model for degree assortativity

```{r sdc_models_assortativity}
as_cm_lm <- lm(atanh(assortativity) ~ space * degseq_type + log(N) + log(m) + log(alpha_n) + degseq_sort, data = sdc)
summary(as_cm_lm)
as_cm_emm <- emmeans(as_cm_lm, ~space + degseq_type)
as_cm_cnt <- contrast(as_cm_emm, method = "pairwise", by = "degseq_type")

sig_df <- tibble(
    degseq_type = c(rep("Poisson", 2), rep("Negative binomial", 2), rep("Power-law", 2)),
    x = c(1.5, 2.5, 1.5, 2.5, 1.5, 2.5),
    y = c(.03, .03, .33, .29, .45, .4),
    label = c("0.001", "0.008", "< 0.001", "< 0.001", "< 0.001", "< 0.001")
)

as_cm_plt <- (as_cm_emm %>%
    tidy %>%
    mutate_at(vars(estimate, conf.low, conf.high), tanh) %>%
    ggplot(aes(x = space, y = estimate, color = degseq_type, group = degseq_type)) +
    geom_line(linetype = 2) +
    geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = .25) +
    geom_point(shape = 16, size = 2) +
    geom_label(data = sig_df, aes(x = x, y = y, label = label, color = degseq_type), size = 3) +
    scale_color_manual(values = ALPHA_COLORS, name = "Degree sequence") +
    scale_y_continuous(limits = c(0, .55), breaks = c(0, .1, .2, .3, .4, .5)) +
    theme(
        axis.text.x = element_text(size = 9),
        legend.position = "none"
    )
) %>%
    add_base_theme +
    xlab("") +
    ylab("Degree assortativity")
```

### Model based comparison of different types of social spaces

```{r sdc_models_plot, fig.width = 8}
ggarrange(cl_cm_plt, as_cm_plt, ncol = 2, common.legend = TRUE, legend = "top")
```
